\documentclass{ximera}
\input{../preamble.tex}

\author{Paul Zachlin \and Anna Davis} \title{Supplementary Exercises for Ch 2} \license{CC-BY 4.0}

\begin{document}

\begin{abstract}
\end{abstract}
\maketitle

\section*{Supplementary Exercises for Ch 2 Systems of Linear Equations}



\begin{problem}\label{prb:possible}
We show in \href{https://ximera.osu.edu/oerlinalg/LinearAlgebra/RRN-0030/main}{Planes in $\RR^3$} that the graph of an equation $ax + by + cz = d$ is a plane in space when not all of $a$, $b$, and $c$ are zero.

\begin{enumerate}
\item By examining the possible positions of planes in space, show that three equations in three variables can have zero, one, or infinitely many solutions.
 %ANSWER  No. If the corresponding planes are parallel and distinct, there is no
%solution. Otherwise they either coincide or have a whole common line of
%solutions, that is, at least one parameter.

\item Can two equations in three variables have a unique solution? Give reasons for your answer.
\end{enumerate}
\end{problem}

\begin{problem}\label{prb:systems_review}
Find all solutions to the following systems of linear equations.

\begin{enumerate}
\item

$\begin{array}{rlrlrlrcr}
	  x_1 & + &  x_2 & + &  x_3 & - &  x_4 & = &  3 \\
	 3x_1 & + & 5x_2 & - & 2x_3 & + &  x_4 & = &  1 \\
	-3x_1 & - & 7x_2 & + & 7x_3 & - & 5x_4 & = &  7 \\
	  x_1 & + & 3x_2 & - & 4x_3 & + & 3x_4 & = & -5 \\
\end{array}$

%ANSWER \hspace*{0em}\newline $x_1 = \frac{1}{10}(-6s - 6t + 16)$, $x_2 = \frac{1}{10}(4s - t + 1)$, $x_3 = s$, $x_4 = t$

\item

$\begin{array}{rlrlrlrcr}
	  x_1 & + &  4x_2 & - &   x_3 & + &   x_4 & = &  2 \\
	 3x_1 & + &  2x_2 & + &   x_3 & + &  2x_4 & = &  5 \\
	  x_1 & - &  6x_2 & + &  3x_3 &   &       & = &  1 \\
	  x_1 & + & 14x_2 & - &  5x_3 & + &  2x_4 & = &  3 \\
\end{array}$

\end{enumerate}

\end{problem}

\begin{problem}\label{prb:no._sol}
In each case find (if possible) conditions on $a$, $b$, and $c$ such that the system has zero, one, or infinitely many solutions.
\begin{enumerate}
\item

$\begin{array}{rlrlrcr}
	  x & + &  2y & - &  4z & = &  4 \\
	 3x & - &   y & + & 13z & = &  2 \\
	 4x & + &   y & + & a^2z & = & a + 3 \\
\end{array}$
%ANSWER  If $a = 1$, no solution. If $a = 2$, $x = 2 - 2t$, $y = -t$, $z = t$. If $a \neq 1$ and $a \neq 2$, the unique          solution is $x = \frac{8 - 5a}{3(a - 1)}$, $y = \frac{-2 - a}{3(a - 1)}$, $z = \frac{a + 2}{3}$

\item

$\begin{array}{rlrlrcr}
	  x & + &   y & + &  3z & = &  a \\
	 ax & + &   y & + &  5z & = &  4 \\
	  x & + &  ay & + &  4z & = &  a
\end{array}$
\end{enumerate}
\end{problem}

\begin{problem}\label{prb:elementaryrowops}
Show that any two rows of a matrix can be interchanged by elementary row transformations of the other two types.

%\begin{supsol}
%$\leftB \begin{array}{c}
%	R_1 \\
%	R_2
%\end{array} \rightB
%\to $ \\ $\leftB \begin{array}{c}
%	R_1 + R_2 \\
%	R_2
%\end{array} \rightB
%\to
%\leftB \begin{array}{c}
%	R_1 + R_2 \\
%	-R_1
%\end{array} \rightB
%\to
%\leftB \begin{array}{c}
%	R_2 \\
%	-R_1
%\end{array} \rightB
%\to
%\leftB \begin{array}{c}
%	R_2 \\
%	R_1
%\end{array} \rightB$
%\end{supsol}
\end{problem}

\begin{problem}\label{prb:rref_nonsingular}
If $ad \neq bc$, show that $\begin{bmatrix}
a & b \\
c & d
\end{bmatrix}$ has reduced row-echelon form $\begin{bmatrix}
 1 & 0 \\
 0 & 1
 \end{bmatrix}$.
\end{problem}

\begin{problem}\label{prb:coeffs}
Find $a$, $b$, and $c$ so that the system
\begin{equation*}
\begin{array}{rlrlrcr}
	 x & + &  ay & + &  cz & = &  0 \\
	bx & + &  cy & - &  3z & = &  1 \\
	ax & + &  2y & + &  bz & = &  5
\end{array}
\end{equation*}
has the solution $x = 3$, $y = -1$, $z = 2$.

$a = \answer{1}$, $b = \answer{2}$, $c = \answer{-1}$

\end{problem}

\begin{problem}\label{prb:complex}
Solve the system
$$
\begin{array}{rlrlrcr}
	 x & + &  2y & + &  2z & = & -3 \\
	2x & + &   y & + &   z & = & -4 \\
	 x & - &   y & + &  iz & = &  i
\end{array}
$$
where $i^{2} = -1$. 
\end{problem}

\begin{problem}\label{prb:real_complex}
Show that the \textit{real} system
\begin{equation*}
\left \{
\begin{array}{rlrlrcr}
	 x & + &   y & + &   z & = &  5 \\
	2x & - &   y & - &   z & = &  1 \\
	-3x& + &  2y & + &  2z & = &  0
\end{array}
\right.
\end{equation*}
has a \textit{complex} solution: $x = 2$, $y = i$, $z = 3 - i$ where $i^{2} = -1$. Explain. What happens when such a real system has a unique solution?

%ANSWER  The (real) solution is $x = 2$, $y = 3 - t$, $z = t$ where $t$ is a parameter. The given complex solution occurs when $t = 3 - i$ is complex. If the real system has a unique solution, that solution is real because the coefficients and constants are all real.

\end{problem}

\begin{problem}\label{prb:pills}
A man is ordered by his doctor to take $5$ units of vitamin A, $13$ units of vitamin B, and $23$ units of vitamin C each day. Three brands of vitamin pills are available, and the number of units of each vitamin per pill are shown in the accompanying table.

$$
\begin{array}{cccc}
\hline
		\textbf{Brand} &  \textbf{A} & \textbf{B} & \textbf{C} \\ \hline
		\textbf{1} & 1 & 2 & 4 \\
		\textbf{2} & 1 & 1 & 3 \\
		\textbf{3} & 0 & 1 & 1 \\ \hline
 \end{array}
$$

\begin{enumerate}
\item Find all combinations of pills that provide exactly the required amount of vitamins (no partial pills allowed).

\item If brands 1, 2, and 3 cost 3 cents, 2 cents, and 5 cents \ per pill, respectively, find the least expensive treatment.

$\answer{5}$ of brand 1, $\answer{0}$ of brand 2, $\answer{3}$ of brand 3

\end{enumerate}
\end{problem}

\begin{problem}\label{prb:tables}
A restaurant owner plans to use $x$ tables seating $4$, $y$ tables seating $6$, and $z$ tables seating $8$, for a total of $20$ tables. When fully occupied, the tables seat $108$ customers. If only half of the $x$ tables, half of the $y$ tables, and one-fourth of the $z$ tables are used, each fully occupied, then $46$ customers will be seated. Find $x$, $y$, and $z$.

\end{problem}

\begin{problem}\label{prb:2x2rref_forms}

\begin{enumerate}
\item Show that a matrix with two rows and two columns that is in reduced row-echelon form must have one of the following forms:
\begin{equation*}
\begin{bmatrix}
1 & 0 \\
0 & 1
\end{bmatrix}
\begin{bmatrix}
0 & 1 \\
0 & 0
\end{bmatrix}
\begin{bmatrix}
0 & 0 \\
0 & 0
\end{bmatrix}
\begin{bmatrix}
1 & * \\
0 & 0
\end{bmatrix}
\end{equation*}

\begin{hint}
 The leading $1$ in the first row must be in column 1 or 2 or not exist.
\end{hint}

\item List the seven reduced row-echelon forms for matrices with two rows and three columns.

\item List the four reduced row-echelon forms for matrices with three rows and two columns.

\end{enumerate}
\end{problem}

\begin{problem}\label{prb:amusement}
An amusement park charges \$$7$ for adults, \$$2$ for youths, and \$$0.50$ for
children. If $150$ people enter and pay a total of \$$100$, find the numbers
of adults, youths, and children.
\begin{hint}
These numbers are nonnegative \textit{integers}.
\end{hint}

\end{problem}

\begin{problem}\label{prb:quadratic}
Solve the following system of equations for $x$ and $y$.

\begin{equation*}
\begin{array}{rlrlrcr}
	  x^2 & + &   xy & - &   y^2 & = &  1 \\
	 2x^2 & - &   xy & + &  3y^2 & = & 13 \\
	  x^2 & + &  3xy & + &  2y^2 & = &  0 \\
\end{array}
\end{equation*}
\begin{hint}
These equations are linear in the new variables $x_{1} = x^{2}$, $x_{2} = xy$, and $x_{3} = y^{2}$.
\end{hint}
\end{problem}





\section*{Text Source} This section was adapted from the second part of Section 8.1 of Keith Nicholson's \href{https://open.umn.edu/opentextbooks/textbooks/linear-algebra-with-applications}{\it Linear Algebra with Applications}. (CC-BY-NC-SA)

W. Keith Nicholson, {\it Linear Algebra with Applications}, Lyryx 2018, Open Edition, p. 415 

%\section*{Example Source}
Example \ref{ex:OrthogDecomp}  was adapted from Example 4.148  of Ken Kuttler's \href{https://open.umn.edu/opentextbooks/textbooks/a-first-course-in-linear-algebra-2017}{\it A First Course in Linear Algebra}. (CC-BY)

Ken Kuttler, {\it  A First Course in Linear Algebra}, Lyryx 2017, Open Edition, p. 249. 

\end{document}