\documentclass{ximera}
\input{../preamble.tex}
\title{Inner Product Spaces} \license{CC BY-NC-SA 4.0}
\begin{document}
\begin{abstract}
 \end{abstract}
\maketitle

\section*{Inner Product Spaces}

We have used the dot product in $\RR^n$ to compute the length of vectors (Corollary \ref{cor:length_via_dotprod}) and also the \href{https://ximera.osu.edu/oerlinalg/LinearAlgebra/VEC-0060/main}{angle between vectors}. The goal of this section is to define an \dfn{inner product} on an arbitrary vector space $V$ over the real numbers. The dot product is an example an inner product for $\RR^n$.

\begin{definition}\label{def:innerproductspace}
An \dfn{inner product} on a real vector space $V$ is a function that assigns a real number $\langle\vec{v}, \vec{w}\rangle$ to every pair $\vec{v}$, $\vec{w}$ of vectors in $V$ in such a way that the following properties are satisfied.

\begin{enumerate}
\item\label{prop:inner_prod_1}  $\langle\vec{v}, \vec{w}\rangle$ \textit{is a real number for all} $\vec{v}$ \textit{and} $\vec{w}$ \textit{in} $V$.

\item\label{prop:inner_prod_2}  $\langle\vec{v}, \vec{w}\rangle = \langle\vec{w}, \vec{v}\rangle$ \textit{for all} $\vec{v}$ \textit{and} $\vec{w}$ \textit{in} $V$.

\item\label{prop:inner_prod_3}  $\langle\vec{v} + \vec{w}, \vec{u}\rangle = \langle\vec{v}, \vec{u}\rangle + \langle\vec{w}, \vec{u}\rangle$ \textit{for all} $\vec{u}$, $\vec{v}$, \textit{and} $\vec{w}$ \textit{in} $V$.

\item\label{prop:inner_prod_4} $\langle r\vec{v}, \vec{w}\rangle = r\langle\vec{v}, \vec{w}\rangle$ \textit{for all} $\vec{v}$ \textit{and} $\vec{w}$ \textit{in} $V$ \textit{and all} $r$ \textit{in} $\RR$.

\item\label{prop:inner_prod_5}  $\langle\vec{v}, \vec{v}\rangle > 0$ \textit{for all} $\vec{v} \neq \vec{0}$ \textit{in} $V$.

\end{enumerate}
\end{definition}
A real vector space $V$ with an inner product $\langle\ , \rangle$ will be called an \dfn{inner product space}.  Note that every subspace of an inner product space is again an inner product space using the same inner product.

\begin{example}\label{exa:030303}
$\RR^n$ is an inner product space with the dot product as inner product:
\begin{equation*}
\langle \vec{v}, \vec{w} \rangle = \vec{v} \dotp \vec{w} \quad \mbox{ for all } \vec{v},  \vec{w} \in \RR^n
\end{equation*}
See Theorem~\ref{th:dotproductproperties}. This is also called the
\dfn{euclidean} inner product, and
$\RR^n$, equipped with the dot product, is called \dfn{euclidean}
$n$-\dfn{space}.
\end{example}

\begin{example}\label{exa:030310}
If $A$ and $B$ are $m \times n$ matrices, define $\langle A, B\rangle = \mbox{tr}(AB^{T})$ where $\mbox{tr}(X)$ is the trace of the square matrix $X$. Show that $\langle\ , \rangle$ is an inner product in $\mathbb{M}_{mn}$.

\begin{explanation}
Property \ref{prop:inner_prod_1} is clear. Since $\mbox{tr}(P) = \mbox{tr}(P^{T})$ for every square matrix $P$, we have Property \ref{prop:inner_prod_2}:
\begin{equation*}
\langle A, B \rangle = \mbox{tr}(AB^T) = \mbox{tr}[(AB^T)^T] = \mbox{tr}(BA^T) = \langle B, A \rangle
\end{equation*}
Next, Property \ref{prop:inner_prod_3} and Property \ref{prop:inner_prod_4} follow because trace is a linear transformation $\mathbb{M}_{mn} \to \RR$ (Practice Problem \ref{ex:10_1_19}). Turning to Property \ref{prop:inner_prod_5}, let $\vec{r}_{1}, \vec{r}_{2}, \dots, \vec{r}_{m}$ denote the rows of the matrix $A$. Then the $(i, j)$-entry of $AA^{T}$ is $\vec{r}_{i} \dotp \vec{r}_{j}$, so
\begin{equation*}
\langle A, A \rangle = \mbox{tr}(AA^T) =
\vec{r}_1 \dotp \vec{r}_1 +
\vec{r}_2 \dotp \vec{r}_2 + \dots +
\vec{r}_m \dotp \vec{r}_m
\end{equation*}
But $\vec{r}_{j} \dotp \vec{r}_{j}$ is the sum of the squares of
the entries of $\vec{r}_{j}$, so this shows that $\langle A, A\rangle$ is the sum of the squares of all $nm$ entries of $A$.  Property \ref{prop:inner_prod_5} follows.
\end{explanation}
\end{example}

The next example is important in analysis.

\begin{example}\label{exa:030334}
Let $\mathcal{F}[a,b]$ be the set of all functions $f:[a,b]\rightarrow\RR$.  Observe that $\mathcal{F}[a,b]$ is a vector space.  Let $\mathcal{C}[a,b]$ be a subset of $\mathcal{F}[a,b]$ consisting of all continuous functions.  Why is $\mathcal{C}[a,b]$ a subspace of $\mathcal{F}[a,b]$?
Show that
\begin{equation*}
\langle f, g \rangle = \int_{a}^{b} f(x)g(x)dx
\end{equation*}
defines an inner product on $\mathcal{C}[a, b]$.

\begin{remark}
    This example (and others later that refer to it) can be omitted with no loss of continuity by students with no calculus background.
\end{remark}

\begin{explanation}
 Property \ref{prop:inner_prod_1} and Property \ref{prop:inner_prod_2} are clear. As to  Property \ref{prop:inner_prod_4},
\begin{equation*}
\langle rf, g \rangle = \int_{a}^{b} rf(x)g(x)dx = r\int_{a}^{b} f(x)g(x)dx =
r\langle f, g \rangle
\end{equation*}
 Property \ref{prop:inner_prod_3} is similar. Finally, theorems of calculus show that $\langle f, f \rangle = \int_{a}^{b} f(x)^2dx \geq 0$ and, if $f$ is continuous, that this is zero if and only if $f$ is the zero function. This gives  Property \ref{prop:inner_prod_5}.
\end{explanation}
\end{example}


If $\vec{v}$ is any vector, then, using  Property \ref{prop:inner_prod_3} of Definition \ref{def:innerproductspace}, we get
\begin{equation*}
\langle \vec{0}, \vec{v} \rangle = \langle \vec{0} + \vec{0}, \vec{v} \rangle =
\langle \vec{0}, \vec{v} \rangle + \langle \vec{0}, \vec{v} \rangle
\end{equation*}
and it follows that the number $\langle\vec{0}, \vec{v}\rangle$ must be zero. This observation is recorded for reference in the following theorem, along with several other properties of inner products. The other proofs are left as Practice Problem \ref{ex:10_1_20}.

\begin{theorem}\label{thm:030346}
Let $\langle\ , \rangle$ be an inner product on a space $V$; let $\vec{v}$, $\vec{u}$, and $\vec{w}$ denote vectors in $V$; and let $r$ denote a real number.

\begin{enumerate}
\item\label{030346a} $\langle \vec{u}, \vec{v} + \vec{w}\rangle =
  \langle\vec{u}, \vec{v}\rangle + \langle\vec{u}, \vec{w}\rangle$

\item\label{030346b} $\langle\vec{v}, r\vec{w}\rangle =  r\langle\vec{v}, \vec{w}\rangle = \langle r\vec{v}, \vec{w}\rangle$

\item\label{030346c} $\langle\vec{v}, \vec{0}\rangle = 0 = \langle\vec{0}, \vec{v} \rangle $

\item\label{030346d} $\langle\vec{v}, \vec{v}\rangle = 0$ if and only if $\vec{v} = \vec{0}$

\end{enumerate}
\end{theorem}

\begin{remark}
    If $r$ is a complex number, then \ref{030346b} must be modified.  See Theorem \ref{th:025575} in \href{https://ximera.osu.edu/oerlinalg/LinearAlgebra/RTH-0050/main}{Complex Matrices}.
\end{remark}

If $\langle\ , \rangle$ is an inner product on a space $V$, then, given $\vec{u}$, $\vec{v}$, and $\vec{w}$ in $V$,
\begin{equation*}
\langle r\vec{u} + s\vec{v}, \vec{w} \rangle = \langle r\vec{u}, \vec{w} \rangle + \langle s\vec{v}, \vec{w} \rangle = r\langle \vec{u}, \vec{w} \rangle + s\langle \vec{v}, \vec{w} \rangle
\end{equation*}
for all $r$ and $s$ in $\RR$ by  Property \ref{prop:inner_prod_3} and Property \ref{prop:inner_prod_4} of Definition \ref{def:innerproductspace}. Moreover, there is nothing special about the fact that there are two terms in the linear combination or that it is in the first component:
\begin{equation*}
\langle r_1\vec{v}_1 + r_2\vec{v}_2 + \dots + r_n\vec{v}_n, \vec{w} \rangle =
r_1\langle \vec{v}_1, \vec{w} \rangle +
r_2\langle \vec{v}_2, \vec{w} \rangle + \dots +
r_n\langle \vec{v}_n, \vec{w} \rangle
\end{equation*}
and
\begin{equation*}
\langle \vec{v}, s_1\vec{w}_1 + s_2\vec{w}_2 + \dots + s_m\vec{w}_m \rangle =
s_1\langle \vec{v}, \vec{w}_1 \rangle +
s_2\langle \vec{v}, \vec{w}_2 \rangle + \dots +
s_m\langle \vec{v}, \vec{w}_m \rangle
\end{equation*}
hold for all $r_{i}$ and $s_{i}$ in $\RR$ and all $\vec{v}$, $\vec{w}$, $\vec{v}_{i}$, and $\vec{w}_{j}$ in $V$. These results are described by saying that inner products ``preserve'' linear combinations. For example,
\begin{align*}
\langle 2\vec{u} - \vec{v}, 3\vec{u} + 2\vec{v} \rangle &=
\langle 2\vec{u}, 3\vec{u} \rangle + \langle 2\vec{u}, 2\vec{v} \rangle + \langle -\vec{v}, 3\vec{u} \rangle + \langle -\vec{v}, 2\vec{v} \rangle \\
&= 6 \langle \vec{u}, \vec{u} \rangle + 4 \langle \vec{u}, \vec{v} \rangle -3 \langle \vec{v}, \vec{u} \rangle - 2 \langle \vec{v}, \vec{v} \rangle \\
&= 6 \langle \vec{u}, \vec{u} \rangle + \langle \vec{u}, \vec{v} \rangle - 2 \langle \vec{v}, \vec{v} \rangle
\end{align*}

If $A$ is a symmetric $n \times n$ matrix and $\vec{x}$ and $\vec{y}$ are columns in $\RR^n$, we regard the $1 \times 1$ matrix $\vec{x}^{T}A\vec{y}$ as a number. If we write
\begin{equation*}
\langle \vec{x}, \vec{y} \rangle = \vec{x}^TA\vec{y} \quad \mbox{ for all columns } \vec{x}, \vec{y} \mbox{ in } \RR^n
\end{equation*}
then  Properties \ref{prop:inner_prod_1} -\ref{prop:inner_prod_4} of Definition \ref{def:innerproductspace} follow from matrix arithmetic (only Property \ref{prop:inner_prod_2} of Definition \ref{def:innerproductspace} requires that $A$ is symmetric).  Property \ref{prop:inner_prod_5} of Definition \ref{def:innerproductspace} reads
\begin{equation*}
\vec{x}^TA \vec{x} > 0 \quad \mbox{ for all columns } \vec{x} \neq \vec{0} \mbox{ in } \RR^n
\end{equation*}
and this condition characterizes the positive definite matrices  (Theorem~\ref{thm:024830}). This proves the first assertion in the next theorem.

\begin{theorem}\label{thm:030372}
If $A$ is any $n \times n$ positive definite matrix, then
\begin{equation*}
\langle \vec{x}, \vec{y} \rangle = \vec{x}^TA\vec{y} \mbox{ for all columns } \vec{x}, \vec{y} \mbox{ in } \RR^n
\end{equation*}
defines an inner product on $\RR^n$, and every inner product on $\RR^n$ arises in this way.
\end{theorem}

\begin{proof}
Given an inner product $\langle\ , \rangle$ on $\RR^n$, let $\{\vec{e}_{1}, \vec{e}_{2}, \dots, \vec{e}_{n}\}$ be the standard basis of $\RR^n$. If $\vec{x} = \displaystyle \sum_{i = 1}^{n} x_i\vec{e}_i$ and $\vec{y} = \displaystyle \sum_{j = 1}^{n} y_j\vec{e}_j$ are two vectors in $\RR^n$, compute $\langle\vec{x}, \vec{y}\rangle$ by adding the inner product of each term $x_{i}\vec{e}_{i}$ to each term $y_{j}\vec{e}_{j}$. The result is a double sum.
\begin{equation*}
\langle \vec{x}, \vec{y} \rangle = \displaystyle \sum_{i = 1}^{n} \sum_{j = 1}^{n} \langle x_i \vec{e}_i, y_j\vec{e}_j \rangle =
\displaystyle \sum_{i = 1}^{n} \sum_{j = 1}^{n} x_i \langle \vec{e}_i, \vec{e}_j \rangle y_j
\end{equation*}
As the reader can verify, this is a matrix product:
\begin{equation*}
\langle \vec{x}, \vec{y} \rangle =
\left[ \begin{array}{cccc}
x_1 & x_2 & \cdots & x_n \\
\end{array} \right]
\left[ \begin{array}{cccc}
\langle \vec{e}_1, \vec{e}_1 \rangle & \langle \vec{e}_1, \vec{e}_2 \rangle & \cdots & \langle \vec{e}_1, \vec{e}_n \rangle \\
\langle \vec{e}_2, \vec{e}_1 \rangle & \langle \vec{e}_2, \vec{e}_2 \rangle & \cdots & \langle \vec{e}_2, \vec{e}_n \rangle \\
\vdots & \vdots & \ddots & \vdots \\
\langle \vec{e}_n, \vec{e}_1 \rangle & \langle \vec{e}_n, \vec{e}_2 \rangle & \cdots & \langle \vec{e}_n, \vec{e}_n \rangle \\
\end{array} \right]
\left[ \begin{array}{c}
	y_1 \\
	y_2 \\
	\vdots \\
	y_n
\end{array} \right]
\end{equation*}
Hence $\langle\vec{x}, \vec{y}\rangle = \vec{x}^{T}A\vec{y}$, where $A$ is the $n \times n$ matrix whose $(i, j)$-entry is $\langle\vec{e}_{i}, \vec{e}_{j} \rangle$. The fact that
\begin{equation*}
\langle\vec{e}_{i}, \vec{e}_{j}\rangle = \langle\vec{e}_{j}, \vec{e}_{i}\rangle\end{equation*}
shows that $A$ is symmetric. Finally, $A$ is positive definite by Theorem~\ref{thm:024830}.
\end{proof}

Thus, just as every linear operator $\RR^n \to \RR^n$ corresponds to an $n \times n$ matrix, every inner product on $\RR^n$ corresponds to a positive definite $n \times n$ matrix. In particular, the dot product corresponds to the identity matrix $I_{n}$.

\begin{remark}
    If we refer to the inner product space $\RR^n$ without specifying the inner product, we mean that the dot product is to be used.
\end{remark}

\begin{example}\label{exa:030413}
Let the inner product $\langle\ , \rangle$ be defined on $\RR^2$ by
\begin{equation*}
\left \langle
\left[ \begin{array}{c}
v_1 \\
v_2
\end{array} \right], \left[ \begin{array}{c}
w_1 \\
w_2
\end{array} \right]
\right \rangle
= 2v_1w_1 - v_1w_2 - v_2w_1 + v_2w_2
\end{equation*}
Find a symmetric $2 \times 2$ matrix $A$ such that $\langle\vec{x}, \vec{y}\rangle = \vec{x}^{T}A\vec{y}$ for all $\vec{x}$, $\vec{y}$ in $\RR^2$.

\begin{explanation}
The $(i, j)$-entry of the matrix $A$ is the coefficient of $v_{i}w_{j}$ in the expression, so
$ A =
\left[ \begin{array}{rr}
2 & -1 \\
-1 & 1
\end{array} \right]$. Incidentally, if
$\vec{x} =
\left[ \begin{array}{r}
x \\
y
\end{array} \right]$, then
\begin{equation*}
\langle \vec{x}, \vec{x} \rangle = 2x^2 - 2xy + y^2 = x^2 +(x - y)^2 \geq 0
\end{equation*}

for all $\vec{x}$, so $\langle\vec{x}, \vec{x}\rangle = 0$ implies
$\vec{x} = \vec{0}$. Hence $\langle\ , \rangle$ is indeed an inner product, so $A$ is positive definite.
\end{explanation}
\end{example}

Let $\langle\ , \rangle$ be an inner product on $\RR^n$ given as in Theorem~\ref{thm:030372} by a positive definite matrix $A$. If $\vec{x} =
\left[ \begin{array}{cccc}
x_1 & x_2 & \cdots & x_n
\end{array} \right]^T $, then $\langle\vec{x}, \vec{x}\rangle = \vec{x}^{T}A\vec{x}$ is an expression in the variables
$x_{1}, x_{2}, \dots, x_{n}$ called a \dfn{quadratic form}. For more on quadratic forms, see Section 8.8 of [Nicholson], pp. 472--482.

\subsection*{Norm and Distance}

\begin{definition}\label{def:030438}
As in $\RR^n$, if $\langle\ , \rangle$ is an inner product on a space $V$, the \dfn{norm}
 $\norm{\vec{v}}$ of a vector $\vec{v}$ in $V$ is defined by
\begin{equation*}
\norm{ \vec{v} } = \sqrt{\langle \vec{v}, \vec{v} \rangle}
\end{equation*}
We define the \dfn{distance} between vectors $\vec{v}$ and $\vec{w}$ in an inner product space $V$ to be
\begin{equation*}
\mbox{d}(\vec{v}, \vec{w}) = \norm{ \vec{v} - \vec{w} }
\end{equation*}
\end{definition}
\begin{remark}
 If the dot product is used in $\RR^n$, the norm $\norm{\vec{x}}$ of a vector $\vec{x}$ is usually called the \dfn{length} of $\vec{x}$.   
\end{remark}


Note that  Property \ref{prop:inner_prod_5} of Definition \ref{def:innerproductspace} guarantees that
$\langle\vec{v}, \vec{v}\rangle \geq 0$, so $\norm{\vec{v}}$ is a real number.

\begin{example}\label{exa:030446}
%NEED FIGURE

The norm of a continuous function $f = f(x)$ in $\mathcal{C}[a, b]$ (with the inner product from Example~\ref{exa:030334}) is given by
\begin{equation*}
\norm{ f } = \sqrt{\int_{a}^{b} f(x)^2dx}
\end{equation*}
Hence $\norm{ f}^{2}$ is the area beneath the graph of $y = f(x)^{2}$ between $x = a$ and $x = b$.
\end{example}

\begin{example}\label{030454}
Show that $\langle\vec{u} + \vec{v}, \vec{u} - \vec{v}\rangle = \norm{\vec{u}}^{2} - \norm{\vec{v}}^{2}$ in any inner product space.

\begin{explanation}
\begin{align*}
\langle \vec{u} + \vec{v}, \vec{u} - \vec{v} \rangle &= \langle \vec{u}, \vec{u} \rangle - \langle \vec{u}, \vec{v} \rangle + \langle \vec{v}, \vec{u} \rangle - \langle \vec{v}, \vec{v} \rangle \\
&= \norm{ \vec{u} }^2 - \langle \vec{u}, \vec{v} \rangle + \langle \vec{u}, \vec{v} \rangle - \norm{ \vec{v} }^2 \\
&= \norm{ \vec{u} }^2 - \norm{ \vec{v} }^2
\end{align*}
\end{explanation}
\end{example}

A vector $\vec{v}$ in an inner product space $V$ is called a \dfn{unit vector} if $\norm{\vec{v}} = 1$. The set of all unit vectors in $V$ is called the \dfn{unit ball} in $V$. For example, if $V = \RR^2$ (with the dot product) and $\vec{v} = (x, y)$, then
\begin{equation*}
\norm{ \vec{v} }^2 = 1 \quad \mbox{ if and only if } \quad x^2 + y^2 = 1
\end{equation*}
Hence the unit ball in $\RR^2$ is the \dfn{unit circle} $x^{2} + y^{2} = 1$ with centre at the origin and radius $1$. However, the shape of the unit ball varies with the choice of inner product.

\begin{example}\label{exa:030469}
%NEED FIGURE


Let $a > 0$ and $b > 0$. If $\vec{v} = (x, y)$ and $\vec{w} = (x_{1}, y_{1})$, define an inner product on $\RR^2$ by
\begin{equation*}
\langle \vec{v}, \vec{w} \rangle = \frac{xx_1}{a^2} + \frac{yy_1}{b^2}
\end{equation*}
The reader can verify (Practice Problem \ref{ex:10_1_5}) that this is indeed an inner product. In this case
\begin{equation*}
\norm{ \vec{v} }^2 = 1 \quad \mbox{ if and only if } \quad \frac{x^2}{a^2} + \frac{y^2}{b^2} = 1
\end{equation*}
so the unit ball is the ellipse shown in the diagram.
\end{example}

Example~\ref{exa:030469} graphically illustrates the fact that norms and distances in an inner product space $V$ vary with the choice of inner product in $V$.

\begin{theorem}\label{030480}
If $\vec{v} \neq \vec{0}$ is any vector in an inner product space $V$, then $\frac{1}{\norm{ \vec{v} }} \vec{v}$ is the unique unit vector that is a positive multiple of $\vec{v}$.
\end{theorem}

The next theorem reveals an important and useful fact about the relationship between norms and inner products, extending the Cauchy inequality for $\RR^n$ (Theorem~\ref{th:CS}).

\begin{theorem}[Cauchy-Schwarz Inequality]\label{030486}
If $\vec{v}$ and $\vec{w}$ are two vectors in an inner product space $V$, then
\begin{equation*}
\langle \vec{v}, \vec{w} \rangle^2 \leq \norm{ \vec{v} }^2 \norm{ \vec{w} }^2
\end{equation*}
Moreover, equality occurs if and only if one of $\vec{v}$ and $\vec{w}$ is a scalar multiple of the other.
\end{theorem}
\begin{remark}
   Hermann Amandus Schwarz (1843--1921) was a German mathematician at the University of Berlin. He had strong geometric intuition, which he applied with great ingenuity to particular problems. A version of the inequality appeared in 1885. 
\end{remark}

\begin{proof}[Proof of Cauchy-Schwarz Inequality]
Write $\norm{\vec{v}} = a$ and $\norm{\vec{w}} = b$. Using Theorem~\ref{thm:030346} we compute:
\begin{equation}\label{eq:thm10_1_4}
\begin{split}
	\norm{ b\vec{v} - a \vec{w} }^2 &= b^2 \norm{ \vec{v} }^2 - 2ab \langle \vec{v}, \vec{w} \rangle + a^2\norm{ \vec{w} }^2 = 2ab(ab - \langle \vec{v}, \vec{w} \rangle) \\
	\norm{ b\vec{v} + a \vec{w} }^2 &= b^2 \norm{ \vec{v} }^2 + 2ab \langle \vec{v}, \vec{w} \rangle + a^2\norm{ \vec{w} }^2 = 2ab(ab + \langle \vec{v}, \vec{w} \rangle) \\
\end{split}
\end{equation}
It follows that $ab - \langle\vec{v}, \vec{w}\rangle \geq 0$ and
$ab + \langle\vec{v}, \vec{w}\rangle \geq 0$, and hence that $-ab \leq \langle\vec{v}, \vec{w}\rangle \leq ab$. But then $| \langle\vec{v}, \vec{w}\rangle | \leq ab = \norm{\vec{v}} \norm{ \vec{w}}$, as desired.

Conversely, if $|\langle \vec{v},  \vec{w}\rangle | =
\norm{\vec{v}} \norm{ \vec{w} } = ab$
then $\langle\vec{v}, \vec{w}\rangle = \pm ab$. Hence (\ref{eq:thm10_1_4}) shows that $b\vec{v} - a\vec{w} = \vec{0}$ or $b\vec{v} + a\vec{w} = \vec{0}$. It follows that one of $\vec{v}$ and $\vec{w}$ is a scalar multiple of the other, even if $a = 0$ or $b = 0$.
\end{proof}

\begin{example}\label{exa:030499}
If $f$ and $g$ are continuous functions on the interval $[a, b]$, then (see Example~\ref{exa:030334})
\begin{equation*}
\left(\int_{a}^{b} f(x)g(x)dx \right) ^2 \leq \int_{a}^{b} f(x)^2 dx \int_{a}^{b} g(x)^2 dx
\end{equation*}
\end{example}

Another famous inequality, the so-called \dfn{triangle inequality} (See \href{https://ximera.osu.edu/oerlinalg/LinearAlgebra/APX-0010/main}{Triangle Inequality} in the Appendix), also comes from the Cauchy-Schwarz inequality. It is included in the following list of basic properties of the norm of a vector.

\begin{theorem}\label{thm:030504}
If $V$ is an inner product space, the norm $\norm{ \dotp }$ has the following properties.

\begin{enumerate}
\item\label{thm:030504a} $\norm{\vec{v}} \geq 0$ for every vector $\vec{v}$ in $V$.

\item\label{thm:030504b} $\norm{\vec{v}} = 0$ if and only if $\vec{v} = \vec{0}$.

\item\label{thm:030504c} $\norm{ r \vec{v}} = |r|\norm{\vec{v}}$ for every $\vec{v}$ in $V$ and every $r$ in $\RR$.

\item\label{thm:030504d} $\norm{\vec{v} + \vec{w}} \leq \norm{\vec{v}} + \norm{\vec{w}}$ for all $\vec{v}$ and $\vec{w}$ in $V$ (\dfn{triangle inequality}).

\end{enumerate}
\end{theorem}

\begin{proof}
Because $\norm{ \vec{v} } = \sqrt{\langle \vec{v}, \vec{v} \rangle}$, properties \ref{thm:030504a} and \ref{thm:030504b} follow immediately from \ref{thm:030346c} and \ref{thm:030346d} of Theorem~\ref{thm:030346}. As to \ref{thm:030504c}, compute
\begin{equation*}
\norm{ r\vec{v} } ^2 = \langle r\vec{v}, r\vec{v} \rangle = r^2\langle \vec{v}, \vec{v} \rangle = r^2\norm{ \vec{v} }^2
\end{equation*}
Hence \ref{thm:030504c} follows by taking positive square roots. Finally, the fact that $\langle\vec{v}, \vec{w}\rangle \leq \norm{\vec{v}}\norm{\vec{w}}$ by the Cauchy-Schwarz inequality gives
\begin{align*}
\norm{ \vec{v} + \vec{w} } ^2 =
\langle \vec{v} + \vec{w}, \vec{v} + \vec{w} \rangle &=
\norm{ \vec{v} } ^2 + 2 \langle \vec{v}, \vec{w} \rangle +
\norm{ \vec{w} } ^2 \\
&\leq \norm{ \vec{v} } ^2 +
2 \norm{ \vec{v} } \norm{ \vec{w} } +
\norm{ \vec{w} } ^2 \\
&= (\norm{ \vec{v} } + \norm{ \vec{w} })^2
\end{align*}
Hence \ref{thm:030504d} follows by taking positive square roots.
\end{proof}

It is worth noting that the usual triangle inequality for absolute values,
\begin{equation*}
| r + s | \leq |r| + |s| \mbox{ for all real numbers } r \mbox{ and } s
\end{equation*}
is a special case of \ref{thm:030504d} where $V = \RR = \RR^1$ and the dot product $\langle r, s \rangle = rs$ is used.

In many calculations in an inner product space, it is required to show that some vector $\vec{v}$ is zero. This is often accomplished most easily by showing that its norm $\norm{\vec{v}}$ is zero. Here is an example.

\begin{example}\label{030528}
Let $\{\vec{v}_{1}, \dots, \vec{v}_{n}\}$ be a spanning set for an inner product space $V$. If $\vec{v}$ in $V$ satisfies $\langle\vec{v}, \vec{v}_{i}\rangle = 0$ for each $i = 1, 2, \dots, n$, show that $\vec{v} = \vec{0}$.

\begin{explanation}
Write $\vec{v} = r_{1}\vec{v}_{1} + \dots + r_{n}\vec{v}_{n}$, $r_{i}$ in $\RR$. To show that $\vec{v} = \vec{0}$, we show that $\norm{\vec{v}}^{2} = \langle\vec{v}, \vec{v}\rangle = 0$. Compute:
\begin{equation*}
\langle \vec{v}, \vec{v} \rangle
= \langle \vec{v}, r_1\vec{v}_1 + \dots + r_n\vec{v}_n \rangle
= r_1\langle \vec{v}, \vec{v}_1 \rangle + \dots + r_n \langle \vec{v}, \vec{v}_n \rangle
= 0
\end{equation*}
by hypothesis, and the result follows.
\end{explanation}
\end{example}

The norm properties in Theorem~\ref{thm:030504} translate to the following properties of distance familiar from geometry. The proof is Practice Problem \ref{ex:10_1_21}.

\begin{theorem}\label{030545}
Let $V$ be an inner product space.

\begin{enumerate}
\item $\mbox{d}(\vec{v}, \vec{w}) \geq 0$ for all $\vec{v}$, $\vec{w}$ in $V$.

\item $\mbox{d}(\vec{v}, \vec{w}) = 0$ if and only if $\vec{v} = \vec{w}$.

\item $\mbox{d}(\vec{v}, \vec{w}) = \mbox{d}(\vec{w}, \vec{v})$ for all $\vec{v}$ and $\vec{w}$ in $V$.

\item $\mbox{d}(\vec{v}, \vec{w}) \leq \mbox{d}(\vec{v}, \vec{u}) + \mbox{d}(\vec{u}, \vec{w})$ for all $\vec{v}$, $\vec{u}$, and $\vec{w}$ in $V$.

\end{enumerate}
\end{theorem}

\section*{Practice Problems}

\begin{problem}\label{prob:inner_prod_1}
In each case, determine which of  Property \ref{prop:inner_prod_1}--Property \ref{prop:inner_prod_5} in Definition \ref{def:innerproductspace} fail to hold.

\begin{enumerate} 
\item $V = \RR^2$, $\left\langle \begin{bmatrix}x_1\\ y_1\end{bmatrix}, \begin{bmatrix}x_2\\ y_2\end{bmatrix} \right\rangle = x_1y_1x_2y_2$

\item $V = \RR^3$, \\$\left\langle \begin{bmatrix}x_1\\ x_2\\ x_3\end{bmatrix}, \begin{bmatrix}y_1\\ y_2\\ y_3\end{bmatrix} \right\rangle = x_1y_1 - x_2y_2 + x_3y_3$

Click the arrow to see the answer.
\begin{expandable}
Property \ref{prop:inner_prod_5} fails.
\end{expandable}

\item $V = \mathbb{C}$, $\langle z, w \rangle = z\overline{w}$, where $\overline{w}$ is complex
conjugation

Click the arrow to see the answer.
\begin{expandable}
Property \ref{prop:inner_prod_1} fails, as sometimes we get a complex number.  However, we will return to this definition of $\langle\ , \rangle$ in \href{https://ximera.osu.edu/oerlinalg/LinearAlgebra/RTH-0050/main}{Complex Matrices} -- see Definition \ref{def:025549}
\end{expandable}

\item $V = \mathbb{P}^3$, $\langle p(x), q(x) \rangle = p(1)q(1)$

Click the arrow to see the answer.
\begin{expandable}
Property \ref{prop:inner_prod_5} fails.
\end{expandable}

\item $V = \mathbb{M}_{22}$, $\langle A, B \rangle = \mbox{det}(AB)$

\item $V = \mathcal{F}[0, 1]$, $\langle f, g \rangle = f(1)g(0) + f(0)g(1)$

Click the arrow to see the answer.
\begin{expandable}
Property \ref{prop:inner_prod_5} fails.
\end{expandable}

\end{enumerate}
\end{problem}

\begin{problem}\label{prob:inner_prod_2}
Let $V$ be an inner product space. If $U \subseteq V$ is a subspace, show that $U$ is an inner product space using the same inner product.

\begin{hint}
 Property \ref{prop:inner_prod_1}--Property \ref{prop:inner_prod_5} hold in $U$ because they hold in $V$.
\end{hint}
\end{problem}

\begin{problem}\label{prob:inner_prod_3}
In each case, find a scalar multiple of $\vec{v}$ that is a unit vector.

\begin{enumerate} 
\item $\vec{v} = f$ in $\mathcal{C}[0, 1]$ where
$f(x) = x^2$  \\ $\langle f, g \rangle \int_{0}^{1} f(x)g(x)dx$

\item $\vec{v} = f$ in $\mathcal{C}[-\pi, \pi]$ where
$f(x) = \cos x$ \\ $\langle f, g \rangle \int_{-\pi}^{\pi} f(x)g(x)dx$

Click the arrow to see the answer.
\begin{expandable}
$\frac{1}{\sqrt{\pi}}f$
\end{expandable}

\item $\vec{v} =
\left[ \begin{array}{r}
1 \\
3
\end{array} \right]$
in $\RR^2$ where $\langle \vec{v}, \vec{w} \rangle = \vec{v}^T
\left[ \begin{array}{rr}
1 & 1 \\
1 & 2
\end{array} \right]
\vec{w}$

\item $ \vec{v} =
\left[ \begin{array}{r}
3 \\
-1
\end{array} \right]$
in $\RR^2$, $\langle \vec{v}, \vec{w} \rangle = \vec{v}^T
\left[ \begin{array}{rr}
1 & -1 \\
-1 & 2
\end{array} \right]
\vec{w}$

Click the arrow to see the answer.
\begin{expandable}
$\frac{1}{\sqrt{17}}
\left[ \begin{array}{r}
3 \\
-1
\end{array} \right]$
\end{expandable}

\end{enumerate}
\end{problem}

\begin{problem}\label{prob:inner_prod_4}
In each case, find the distance between $\vec{u}$ and $\vec{v}$.

\begin{enumerate} 
\item $\vec{u} = \begin{bmatrix}3\\ -1\\ 2\\ 0\end{bmatrix}$, $\vec{v} = \begin{bmatrix}1\\ 1\\ 1\\ 3\end{bmatrix};
\langle \vec{u}, \vec{v} \rangle = \vec{u} \dotp \vec{v}$

\item $\vec{u} = \begin{bmatrix}1\\  2\\ -1\\ 2\end{bmatrix}$, $\vec{v} = \begin{bmatrix}2\\ 1\\ -1\\ 3\end{bmatrix};
\langle \vec{u}, \vec{v} \rangle = \vec{u} \dotp \vec{v}$

Click the arrow to see the answer.
\begin{expandable}
$\sqrt{3}$
\end{expandable}

\item $\vec{u} = f$, $\vec{v} = g $ in $\mathcal{C}[0, 1]$ where $f(x) = x^2 $ and $g(x) = 1 - x$; $\langle f, g \rangle = \int_{0}^{1} f(x)g(x)dx$

\item $\vec{u} = f$, $\vec{v} = g $ in $\mathcal{C}[-\pi, \pi]$ where $f(x) = 1$ and $g(x) = \cos x$; $\langle f, g \rangle = \int_{-\pi}^{\pi} f(x)g(x)dx$

Click the arrow to see the answer.
\begin{expandable}
$\sqrt{3\pi}$
\end{expandable}

\end{enumerate}
\end{problem}

\begin{problem}\label{ex:10_1_5}
Let $a_{1}, a_{2}, \dots, a_{n}$ be positive numbers. Given $\vec{v} = \begin{bmatrix}v_{1}\\ v_{2}\\ \vdots\\ v_{n}\end{bmatrix}$ and $\vec{w} = \begin{bmatrix}w_{1}\\ w_{2}\\ \vdots\\ w_{n}\end{bmatrix}$, define $\langle\vec{v}, \vec{w}\rangle = a_{1}v_{1}w_{1} + \dots + a_{n}v_{n}w_{n}$. Show that this is an inner product on $\RR^n$.
\end{problem}

\begin{problem}\label{prob:inner_prod_6}
If $\{\vec{b}_{1}, \dots, \vec{b}_{n}\}$ is a basis of $V$ and if
$\vec{v} = v_1\vec{b}_1 + \dots + v_n\vec{b}_n$ and
$\vec{w} = w_1\vec{b}_1 + \dots + w_n\vec{b}_n$ are vectors in $V$, define
\begin{equation*}
\langle \vec{v}, \vec{w} \rangle = v_1w_1 + \dots + v_nw_n .
\end{equation*}
Show that this is an inner product on $V$.
\end{problem}


\begin{problem}\label{prob:inner_prod_9}
Let $\mbox{re}(z)$ denote the real part of the complex number
$z$. Show that $\langle\ , \rangle$ is an inner product on $\mathbb{C}$ if $\langle\vec{z}, \vec{w}\rangle = \mbox{re}(z\overline{w})$.
\end{problem}

\begin{problem}\label{prob:inner_prod_10}
If $T : V \to V$ is an isomorphism of the inner product space $V$, show that
\begin{equation*}
\langle \vec{v}, \vec{w} \rangle_1 = \langle T(\vec{v}), T(\vec{w}) \rangle
\end{equation*}
defines a new inner product $\langle\ , \rangle_{1}$ on $V$.
\end{problem}

\begin{problem}\label{prob:inner_prod_11}
Show that every inner product $\langle\ , \rangle$ on $\RR^n$ has the form $\langle\vec{x}, \vec{y}\rangle = (U\vec{x}) \dotp (U\vec{y})$ for some upper triangular matrix $U$ with positive diagonal entries. 
\begin{hint}
    Theorem~\ref{thm:024907}
\end{hint}
\end{problem}

\begin{problem}\label{prob:inner_prod_12}
In each case, show that $\langle\vec{v}, \vec{w}\rangle = \vec{v}^{T}A\vec{w}$ defines an inner product on $\RR^2$ and hence show that $A$ is positive definite.
\begin{enumerate}
\item
$A =
\left[ \begin{array}{rr}
2 & 1 \\
1 & 1
\end{array} \right]$

\item
$A =
\left[ \begin{array}{rr}
5 & -3 \\
-3 & 2
\end{array} \right]$

Click the arrow to see the answer.
\begin{expandable}
$  \langle \vec{v}, \vec{v} \rangle = 5v_1^2 - 6v_1v_2 + 2v_2^2 =
\frac{1}{5}[(5v_1 - 3v_2)^2 + v_2^2] $
\end{expandable}

\item
$A =
\left[ \begin{array}{rr}
3 & 2 \\
2 & 3
\end{array} \right]$

\item
$A =
\left[ \begin{array}{rr}
3 & 4 \\
4 & 6
\end{array} \right]$
\end{enumerate}

Click the arrow to see the answer.
\begin{expandable}
$ \langle \vec{v}, \vec{v} \rangle = 3v_1^2 + 8v_1v_2 + 6v_2^2 =
\frac{1}{3}[(3v_1 + 4v_2)^2 + 2v_2^2] $
\end{expandable}
\end{problem}

\begin{problem}\label{prob:inner_prod_13}
In each case, find a symmetric matrix $A$ such that $\langle\vec{v}, \vec{w}\rangle = \vec{v}^{T}A\vec{w}$.


\begin{enumerate} 
\item
$  \left\langle
\left[ \begin{array}{r}
v_1 \\
v_2
\end{array} \right], \left[ \begin{array}{r}
w_1 \\
w_2
\end{array} \right]
\right\rangle
= v_1w_1 + 2v_1w_2 + 2v_2w_1 + 5v_2w_2$

\item
$\left\langle
\left[ \begin{array}{r}
v_1 \\
v_2
\end{array} \right], \left[ \begin{array}{r}
w_1 \\
w_2
\end{array} \right]
\right\rangle
= v_1w_1 - v_1w_2 - v_2w_1 + 2v_2w_2$

Click the arrow to see the answer.
\begin{expandable}
$\left[ \begin{array}{rr}
1 & -2 \\
-2 & 1
\end{array} \right]$
\end{expandable}

\item
$\left\langle
\left[ \begin{array}{r}
v_1 \\
v_2 \\
v_3
\end{array} \right], \left[ \begin{array}{r}
w_1 \\
w_2 \\
w_3
\end{array} \right]
\right\rangle
= 2v_1w_1 + v_2w_2 + v_3w_3 - v_1w_2 \\ -v_2w_1 + v_2w_3 + v_3w_2$

\item
$\left\langle
\left[ \begin{array}{r}
v_1 \\
v_2 \\
v_3
\end{array} \right], \left[ \begin{array}{r}
w_1 \\
w_2 \\
w_3
\end{array} \right]
\right\rangle
= v_1w_1 + 2v_2w_2 + 5v_3w_3 \\ - 2v_1w_3 - 2v_3w_1$

Click the arrow to see the answer.
\begin{expandable}
$\left[ \begin{array}{rrr}
1 & 0 & -2 \\
0 & 2 & 0 \\
-2 & 0 & 5
\end{array} \right]$
\end{expandable}

\end{enumerate}
\end{problem}

\begin{problem}\label{prob:inner_prod_14}
If $A$ is symmetric and $\vec{x}^{T}A\vec{x} = 0$ for all columns $\vec{x}$ in $\RR^n$, show that $A = 0$. 

\begin{hint}
Consider
$\langle \vec{x} + \vec{y}, \vec{x} + \vec{y} \rangle$ where $\langle \vec{x}, \vec{y} \rangle = \vec{x}^TA\vec{y}$.
\end{hint}

Click the arrow to see the answer.
\begin{expandable}
By the condition, $\langle \vec{x}, \vec{y} \rangle = \frac{1}{2} \langle \vec{x} + \vec{y}, \vec{x} + \vec{y} \rangle = 0$ for all $\vec{x}$, $\vec{y}$. Let $\vec{e}_{i}$ denote column $i$ of $I$. If $A = \left[ a_{ij} \right]$, then $a_{ij} = \vec{e}_{i}^{T}A\vec{e}_{j} = \{\vec{e}_{i}, \vec{e}_{j}\} = 0$ for all $i$ and $j$.
\end{expandable}
\end{problem}

\begin{problem}\label{prob:inner_prod_15}
Show that the sum of two inner products on $V$ is again an inner product.
\end{problem}

\begin{problem} \label{ex:10_1_16}
Let $ \norm{ \vec{u} } = 1$, $\norm{ \vec{v} } = 2$, $\norm{ \vec{w} } = \sqrt{3} $, $\langle \vec{u}, \vec{v} \rangle = -1$, $\langle\vec{u}, \vec{w}\rangle = 0$ and $\langle\vec{v}, \vec{w}\rangle = 3$. Compute:
\begin{enumerate}
\item $\langle \vec{v} + \vec{w}, 2\vec{u} - \vec{v} \rangle$
\item $\langle \vec{u} - 2 \vec{v} - \vec{w}, 3\vec{w} - \vec{v} \rangle$

$\answer{-15}$
\end{enumerate}

\end{problem}

\begin{problem}\label{prob:inner_prod_17}
Given the data in Practice Problem \ref{ex:10_1_16}, show that $\vec{u} + \vec{v} = \vec{w}$.
\end{problem}

\begin{problem}\label{prob:inner_prod_18}
Show that no vectors exist such that $\norm{\vec{u}} = 1$, $\norm{\vec{v}} = 2$, and $\langle\vec{u}, \vec{v}\rangle = -3$.
\end{problem}

\begin{problem} \label{ex:10_1_19}
Complete Example~\ref{exa:030310}.
\end{problem}

\begin{problem} \label{ex:10_1_20}
Prove Theorem~\ref{thm:030346}.

\begin{hint}
\begin{enumerate}
\item Using Property \ref{prop:inner_prod_2}:
$ \langle \vec{u}, \vec{v} + \vec{w} \rangle =
\langle \vec{v} + \vec{w}, \vec{u} \rangle =
\langle \vec{v}, \vec{u} \rangle + \langle \vec{w}, \vec{u} \rangle =
\langle \vec{u}, \vec{v} \rangle + \langle \vec{u}, \vec{w} \rangle$.

\item Using Property \ref{prop:inner_prod_2} and Property \ref{prop:inner_prod_4}:
$\langle \vec{v}, r\vec{w} \rangle =
\langle r\vec{w}, \vec{v} \rangle =
r \langle \vec{w}, \vec{v} \rangle =
r \langle \vec{v}, \vec{w} \rangle$.

\item Using Property \ref{prop:inner_prod_3}:
$\langle \vec{0}, \vec{v} \rangle =
\langle \vec{0} + \vec{0}, \vec{v} \rangle =
\langle \vec{0}, \vec{v} \rangle + \langle \vec{0}, \vec{v} \rangle
$, so $ \langle \vec{0}, \vec{v} \rangle = 0$. The rest is Property \ref{prop:inner_prod_2}.

\item Assume that $\langle \vec{v}, \vec{v} \rangle = 0$. If $\vec{v} \neq \vec{0}$ this contradicts Property \ref{prop:inner_prod_5}, so $\vec{v} = \vec{0}$. Conversely, if $\vec{v} = \vec{0}$, then $\langle \vec{v}, \vec{v} \rangle = 0$ by Part 3 of this theorem.
\end{enumerate}
\end{hint}
\end{problem}

\begin{problem} \label{ex:10_1_21}
Prove Theorem~\ref{thm:030346}.
\end{problem}

\begin{problem}\label{prob:inner_prod_22}
Let $\vec{u}$ and $\vec{v}$ be vectors in an inner product space $V$.

\begin{enumerate} 
\item Expand $\langle2\vec{u} - 7\vec{v}, 3\vec{u} + 5\vec{v} \rangle$.

\item Expand $\langle3\vec{u} - 4\vec{v}, 5\vec{u} + \vec{v} \rangle$.

Click the arrow to see the answer.
\begin{expandable}
$15\norm{\vec{u}}^{2} - 17 \langle \vec{u}, \vec{v} \rangle - 4\norm{\vec{v}}^{2}$
\end{expandable}

\item Show that $\norm{ \vec{u} + \vec{v} } ^2 = \norm{ \vec{u} } ^2 + 2 \langle \vec{u}, \vec{v} \rangle + \norm{ \vec{v} } ^2 $.

\item Show that $\norm{ \vec{u} - \vec{v} } ^2 = \norm{ \vec{u} } ^2 - 2 \langle \vec{u}, \vec{v} \rangle +
\norm{ \vec{v} } ^2$.

Click the arrow to see the answer.
\begin{expandable}
$\norm{\vec{u} + \vec{v}}^{2} = \langle \vec{u} + \vec{v}, \vec{u} + \vec{v} \rangle = \norm{\vec{u}}^{2} + 2\langle \vec{u}, \vec{v}\rangle + \norm{\vec{v}}^{2}$
\end{expandable}
\end{enumerate}

\end{problem}

\begin{problem}\label{prob:inner_prod_23}
Show that
\begin{equation*}
\norm{ \vec{v} } ^2 +
\norm{ \vec{w} } ^2 = \frac{1}{2} \{
\norm{ \vec{v} + \vec{w} } ^2 +
\norm{ \vec{v} - \vec{w} } ^2\}
\end{equation*}
for any $\vec{v}$ and $\vec{w}$ in an inner product space.
\end{problem}

\begin{problem}\label{prob:inner_prod_24}
Let $\langle\ , \rangle$ be an inner product on a vector space $V$. Show that the corresponding distance function is translation invariant. That is, show that \newline $\mbox{d}(\vec{v}, \vec{w}) = \mbox{d}(\vec{v} + \vec{u}, \vec{w} + \vec{u})$ for all $\vec{v}$, $\vec{w}$, and $\vec{u}$ in $V$.
\end{problem}

\begin{problem}\label{prob:inner_prod_25}
\begin{enumerate} 
\item Show that $\langle \vec{u}, \vec{v} \rangle = \frac{1}{4}[\norm{ \vec{u} + \vec{v} } ^2 - \norm{ \vec{u} - \vec{v} } ^2]$ for all $\vec{u}$, $\vec{v}$ in an inner product space $V$.

\item If $\langle\ , \rangle$ and $\langle\ , \rangle^\prime$ are two inner products on $V$ that have equal associated norm functions, show that $\langle\vec{u}, \vec{v}\rangle = \langle\vec{u}, \vec{v}\rangle^\prime$ holds for all $\vec{u}$ and $\vec{v}$.

\end{enumerate}
\end{problem}

\begin{problem}\label{prob:inner_prod_26}
Let $\vec{v}$ denote a vector in an inner product space $V$.

\begin{enumerate} 
\item Show that $W = \{\vec{w} \mid \vec{w} \mbox{ in } V, \langle\vec{v}, \vec{w} = 0\}$ is a subspace of $V$.

\item Let $W$ be as in (a). If $V = \RR^3$ with the dot product, and if $\vec{v} = \begin{bmatrix}1\\ -1\\ 2\end{bmatrix}$, find a basis for $W$.

Click the arrow to see the answer.
\begin{expandable}
$\left\{\begin{bmatrix}1\\ 1\\ 0\end{bmatrix}, \begin{bmatrix}0\\ 2\\ 1\end{bmatrix}\right\}$
\end{expandable}
\end{enumerate}
\end{problem}

\begin{problem} \label{ex:10_1_27}
Given vectors $\vec{w}_{1}, \vec{w}_{2}, \dots, \vec{w}_{n}$ and $\vec{v}$, assume that $\langle\vec{v}, \vec{w}_{i}\rangle = 0$ for each $i$. Show that $\langle\vec{v}, \vec{w}\rangle = 0$ for all $\vec{w}$ in $\mbox{span}\{\vec{w}_{1}, \vec{w}_{2}, \dots, \vec{w}_{n}\}$.
\end{problem}

\begin{problem}\label{prob:inner_prod_28}
If $V = \mbox{span}\{\vec{v}_{1}, \vec{v}_{2}, \dots, \vec{v}_{n}\}$ and $\langle\vec{v}, \vec{v}_{i}\rangle = \langle\vec{w}, \vec{v}_i\rangle$ holds for each $i$. Show that $\vec{v} = \vec{w}$.

\begin{hint}
$\langle \vec{v} - \vec{w}, \vec{v}_{i} \rangle = \langle \vec{v}, \vec{v}_{i} \rangle - \langle \vec{w}, \vec{v}_{i} \rangle = 0$ for each $i$, so $\vec{v} = \vec{w}$ by Practice Problem \ref{ex:10_1_27}.
\end{hint}
\end{problem}

\begin{problem}\label{prob:inner_prod_29}
Use the Cauchy-Schwarz inequality in an inner product space to show that:

\begin{enumerate} 
\item If $\norm{\vec{u}} \leq 1$, then $\langle\vec{u}, \vec{v}\rangle^{2} \leq \norm{\vec{v}}^{2}$ for all $\vec{v}$ in $V$.

\item $(x \cos \theta + y \sin \theta)^{2} \leq x^{2} + y^{2}$ for all real $x$, $y$, and $\theta$.

\begin{hint}
If $\vec{u} = (\cos \theta, \sin \theta)$ in $\RR^2$ (with the dot product) then $\norm{\vec{u}} = 1$. Use (a) with $\vec{v} = \begin{bmatrix}x\\ y\end{bmatrix}$.
\end{hint}

\item $\norm{ r_1\vec{v}_1 + \dots + r_n\vec{v}_n } ^2 \leq [r_1 \norm{ \vec{v}_1 } + \dots + r_n \norm{ \vec{v}_n } ]^2$
for all vectors $\vec{v}_{i}$, and all $r_{i} > 0$ in $\RR$.

\end{enumerate}

\end{problem}

\begin{problem}\label{prob:inner_prod_30}
If $A$ is a $2 \times n$ matrix, let $\vec{u}$ and $\vec{v}$ denote the rows of $A$.

\begin{enumerate} 
\item Show that
$AA^T = \left[ \begin{array}{rr}
\norm{ \vec{u} } ^2 & \vec{u} \dotp \vec{v} \\
\vec{u} \dotp \vec{v} & \norm{ \vec{v} } ^2
\end{array} \right]$.

\item Show that $\mbox{det}(AA^{T}) \geq 0$.

\end{enumerate}
\end{problem}

\begin{problem} \label{ex:10_1_31}
\begin{enumerate} 
\item If $\vec{v}$ and $\vec{w}$ are nonzero vectors in an inner product space $V$, show that
$-1 \leq \frac{\langle \vec{v}, \vec{w} \rangle}{\norm{ \vec{v} } \norm{ \vec{w} }} \leq 1$, and hence that a unique angle $\theta$ exists such that \newline $\frac{\langle \vec{v}, \vec{w} \rangle}{\norm{ \vec{v} } \norm{ \vec{w} }} = \cos \theta$ and $0 \leq \theta \leq \pi$. This angle $\theta$ is called the
\dfn{angle between} $\vec{v}$ and $\vec{w}$.

\item Find the angle between $\vec{v} = \begin{bmatrix}1\\ 2\\ -1\\ 1\\ 3\end{bmatrix}$ and $\vec{w} = \begin{bmatrix}2\\ 1\\ 0\\ 2\\ 0\end{bmatrix}$ in $\RR^5$ with the dot product.

\item If $\theta$ is the angle between $\vec{v}$ and $\vec{w}$, show that the \dfn{law of cosines} is valid:
\begin{equation*}
\norm{ \vec{v} - \vec{w} } = \norm{ \vec{v} } ^2 + \norm{ \vec{w} } ^2 - 2\norm{ \vec{v} } \norm{ \vec{w} } \cos \theta.
\end{equation*}
\end{enumerate}
\end{problem}

\begin{problem}\label{prob:inner_prod_32}
If $V = \RR^2$, define $\norm{\begin{bmatrix}x\\ y\end{bmatrix}} = |x| + |y|$.

\begin{enumerate} 
\item Show that $\norm{\dotp}$ satisfies the conditions in Theorem~\ref{thm:030504}.

\item Show that $\norm{\dotp}$ does not arise from an inner product on $\RR^2$ given by a matrix $A$. 
\begin{hint}
    If it did, use Theorem~\ref{thm:030372} to find numbers $a$, $b$, and $c$ such that $\norm{\begin{bmatrix}x\\ y\end{bmatrix}}^{2} = ax^{2} + bxy + cy^{2}$ for all $x$ and $y$.
\end{hint}

\end{enumerate}
\end{problem}

\section*{Text Source} This section was adapted from Section 10.1 of Keith Nicholson's \href{https://open.umn.edu/opentextbooks/textbooks/linear-algebra-with-applications}{\it Linear Algebra with Applications}. (CC-BY-NC-SA)

W. Keith Nicholson, {\it Linear Algebra with Applications}, Lyryx 2018, Open Edition, pp. 521--530.

\end{document}